Investigation of thermal reactionomes reveals differing responses to thermal

### 2014-09-25 doc/ApTranscriptome_manuscript.Rmd
updates to ms

---
title: Investigation of thermal reactionomes reveals stress tolerance vs. resistance in
  warm and cold-climate ant species
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

@@

To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions. First, species from more thermally-stressful environments will rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011]. To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
options(scipen = 1, digits = 3)
load("ms_20140924.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
```

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate. Of these, xxx had a non-linear relationship with temperature that would not have been detected with a standard differential expression approach. At a minimum `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. ~~However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]. Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.~~

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. **Low** The total number of genes for two categories (*A. picea* **Low** - *A. carolinensis* **Low**; *A. picea* **NotResp** - *A. carolinensis* **Bimodal**) is labeled in two boxes for perspective."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

@@

In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A. carolinensis*, the total number of thermally-responsive genes did not differ between the species (...`r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])`...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ... ). Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A. carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A. picea* (Fig. 1).

@@

```{r expression_change, echo=FALSE}
cont.tab <- structure(unlist(as.vector(tt2)), .Dim = c(5L, 5L), .Dimnames = structure(list(
    c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", 
    "Intermediate", "High", "Bimodal", "NotResp")), .Names = c("Ac", 
"Ap")))

# marginal sums and total
rs <- rowSums(cont.tab)
cs <- colSums(cont.tab)
n <- sum(cont.tab)

# grand mean for each class
gm <- (rs + cs) / sum(cont.tab * 2)

# calculate observed values for each cell using overall mean for each expression type
Ec <- outer(gm, gm, "*") *sum(cont.tab)

# get deviations of observed from expected
Ec.dev <- cont.tab - Ec

# overall table P-value
Ec.chistat <- sum((cont.tab-Ec)^2 / Ec)
# P-value
pchisq(Ec.chistat, df=(nrow(cont.tab)-1) * (ncol(cont.tab)-1), lower.tail = FALSE)

# overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
df = (ncol(cont.tab) - 1) * (nrow(cont.tab) - 1)
X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)

# posthoc result for each cell
Ec.cells <- sign(cont.tab - Ec) * (cont.tab - Ec)^2 / Ec
Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
                                                            ifelse(x > X2.df, "+", "-")))

# plot
qplot(x=Ap, y=Ac, data=melt(Ec.cells), fill=value, geom="tile") + 
  scale_fill_gradient2(limits=c(-600, 600)) 
```

@@

We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for genes that had *Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

In addition, we examined the magnitude of upregulation...
...overall mean level of expression...

Further, we compared the mean expression of genes that were not thermally-responsive in one species to the expression level near optimum temperatures in the other species. Finding that mean expression of the non-responsive genes than the optimum genes would be indicative of adaptative basal expression....

@@

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the ... heat shock proteins in the transcriptome, ... were upregulated to cold or hot temperatures in *A. picea* while only ... were upregulated in *A. carolinensis*.

@@

By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. We found that ...% of genes have non-linear responses to temperature.

@@

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Fig. 3). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

>>>>

---
title: Investigation of thermal reactionomes reveals differing responses to thermal
  extremes in warm and cold-climate ant species
author: "null"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

**Authors**

John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4},
Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}

\vspace{10 mm}

**Affilitiations**

\textsuperscript{1} Department of Biology, University of Vermont \newline
\textsuperscript{2} Vermont Genetics Network, University of Vermont \newline
\textsuperscript{3} Department of Biological Sciences, North Carolina State University \newline
\textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions. First, species from more thermally-stressful environments will rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter.

Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-25.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
```

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate. Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. This result is the same if we focus on only the transcripts that have a BLAST hit. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e. the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other."}

rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
```

@@

In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A. carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ... ). Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A. carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A. picea* (Fig. 1).

@@

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of genes based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}
#tt2

# marginal sums and total
rs <- rowSums(tt2)
cs <- colSums(tt2)
n <- sum(tt2)

# grand mean for each class
gm <- (rs + cs) / sum(tt2 * 2)

# calculate observed values for each cell using overall mean for each expression type
Ec <- outer(gm, gm, "*") *sum(tt2)

# get deviations of observed from expected
Ec.dev <- tt2 - Ec

# overall table P-value
Ec.chistat <- sum((tt2-Ec)^2 / Ec)
# P-value
pchisq(Ec.chistat, df=(nrow(tt2)-1) * (ncol(tt2)-1), lower.tail = FALSE)

# overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
df = (ncol(tt2) - 1) * (nrow(tt2) - 1)
X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)

# posthoc result for each cell
Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
                                                            ifelse(x > X2.df, "+", "-")))

md <- melt(Ec.cells)

# plot
mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) 
mh_plot
```

@@

We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, in multiple comparisons. First, we predicted that gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea* should be higher in the opposite species (e.g. greater basal expression but lower inducibility). Instead, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for genes that had **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

Second, for the genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally-responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1), we predicted that mean expression level of the non-responsive *A. carolinensis* genes would be greater than the expression level near the optimum temperature (19.5°C) of the responsive *A. picea* genes. Again, we found no support for genetic assimilation as the non-responsive genes had lower mean expression levels.

@@

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.

@@

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.
